\chapter{Node Architecture}
\label{cha:node-architecture}


\section{Overview}

This chapter describes the architecture of INET host and router models.

Hosts and routers in the INET Framework are OMNeT++ compound modules that
are composed of the following ingredients:

\begin{itemize}

\item \tbf{Interface Table (\nedtype{InterfaceTable})}. This module
contains the table of network interfaces (eth0, wlan0, etc) in the host.
Interfaces are registered dynamically during the initialization phase by
modules that represent network interface cards (NICs). Other modules access
interface information via a C++ class interface.

\item \tbf{Routing Table (\nedtype{RoutingTable})}. This module contains
the IPv4 routing table. It is also accessed from other via a C++ interface.
The interface contains member functions for adding, removing, enumerating
and looking up routes, and finding the best matching route for a given
destination IP address. The IP module calls uses the latter function for
routing packets, and routing protocols such as OSPF or BGP call the route
manipulation methods to add and manage discovered routes.

\item \tbf{Notification Board (\nedtype{NotificationBoard})}. This module
makes it possible for several modules to communicate in a publish-subscribe
manner. Notifications include change notifications (``routing table
changed'') and state changes (``radio became idle'').

\item \tbf{Mobility module}. In simulations involving node mobility, this
module is responsible for moving around the node in the simulated
``playground.'' A mobility module is also needed for wireless simulations
even if the node is stationery, because the mobility module stores the
node's location, needed to compute wireless transmissions. Different
mobility models (Random Walk, etc.) are supported via different module
types, and many host models define their mobility submodules with
parametric type so that the mobility model can be changed in the
configuration (\ttt{"mobility: <mobilityType> like IMobility"}).

\item \tbf{NICs}. Network interfaces are usually compound modules
themselves, composed of a queue and a MAC module. Examples are
\nedtype{PPPInterface}, \nedtype{EthernetInterface}, and WLAN interfaces
such as \nedtype{Ieee80211NicSTA}. The queue submodule stores packets
waiting for transmission, and it is usually defined as having parametric
type as it has multiple implementations to accomodate different needs
(\nedtype{DropTailQueue}, \nedtype{REDQueue}, \nedtype{DropTailQoSQueue},
etc.) Most MACs also have an internal queue to allow operation without an
external queue module, the advantage being smaller overhead. The NIC's
entry in the host's \nedtype{InterfaceTable} is usually registered by the
MAC module at the beginning of the simulation.

\item \tbf{Network layer}. Modules that represent protocols of the network
layer are usually grouped into a compound module: \nedtype{NetworkLayer}
for IP, and \nedtype{NetworkLayer6} for IPv6. \nedtype{NetworkLayer}
contains the modules \nedtype{IP}, \nedtype{ARP}, \nedtype{ICMP} and
\nedtype{ErrorHandling}. The \nedtype{IP} module performs IP
encapsulation/decapsulation and routing of datagrams; for the latter it
accesses the C++ function call interface of the \nedtype{RoutingTable}.
Packet forwarding can be turned on/off via a module parameter. The
\nedtype{ARP} module is put into the path of packets leaving the network
layer towards the NICs, and performs address resolution for interfaces that
need it (e.g. Ethernet). \nedtype{ICMP} deals with sending and receiving
ICMP packets. The \nedtype{ErrorHandling} module receives and logs ICMP
error replies. The IPv6 network layer, \nedtype{NetworkLayer6} contains the
modules \nedtype{IPv6}, \nedtype{ICMPv6}, \nedtype{IPv6NeighbourDiscovery}
and \nedtype{IPv6ErrorHandling}.

\item \tbf{Transport layer protocols}. Transport protocols are represented
by modules connected to the network layer; currently TCP, UDP and SCTP are
supported. TCP has several implementations: \nedtype{TCP} is the OMNeT++
native implementation; the \nedtype{TCP\_lwip} module wraps the lwIP TCP
stack [TODO]; and the \nedtype{TCP\_NSC} module wraps the Network
Simulation Cradle library [TODO]. For this reason, the \ttt{tcp} submodule
is usually defined with a parametric submodule type (\ttt{"tcp: <tcpType>
like ITCP"}).  UDP and SCTP are implemented by the \nedtype{UDP} and
\nedtype{SCTP} modules, respectively.

\item \sloppypar \tbf{Applications}. Application modules typically connect to TCP
and/or UDP, and model the user behavior as well as the application program
(e.g. browser) and application layer protocol (e.g. HTTP). For convenience,
\nedtype{StandardHost} supports any number of UDP, TCP and SCTP
applications, their types being parametric (\ttt{"tcpApp[numTcpApps]:
<tcpAppType> like TCPApp; udpApp[numUdpApps]: <udpAppType> like
UDPApp; ..."}). This way the user can configure applications entirely from
\ttt{omnetpp.ini}, and does not need to write a new NED file every time
different applications are needed in a host model. Application modules
are typically not present in router models.

\item \tbf{Routing protocols}. For routers; routing protocol above TCP and/or UDP

\item \tbf{MPLS modules}. Additional modules if IPv6/MIPv6 or MPLS is used

\end{itemize}

For routers: routing protocol above TCP and/or UDP

Additional modules if IPv6/MIPv6 or MPLS is used

Battery if sensor network

The predefined NED types \nedtype{StandardHost} and \nedtype{Router} are
only one possible way of building host/router models; for specific
simulations it is a perfectly valid approach to create custom node models.

TODO What is needed when...  (I.e. when can InterfaceTable be omitted, RoutingTable, queue, etc)


\section{Addresses}

The INET Framework uses a number of C++ classes to represent various
addresses in the network. These classes support initialization and
assignment from binary and string representation of the address, and
accessing the address in both forms. (Storage is in binary form). They also
support the "unspecified" special value (and the \ffunc{isUnspecified()}
method) that corresponds to the all-zeros address.

\begin{itemize}
  \item \cppclass{MACAddress} represents a 48-bit IEEE 802 MAC address. The
    textual notation it understands and produces is hex string.

  \item \cppclass{IPAddress} represents a 32-bit IPv4 address. It can parse
    and produce textual representations in the "dotted decimal" syntax.

  \item \cppclass{IPv6Address} represents a 128-bit IPv6 address. It can parse
    and produce address strings in the canonical (RFC 3513) syntax.

  \item \cppclass{IPvXAddress} is conceptually a union of a \cppclass{IPAddress}
    and \cppclass{IPv6Address}: an instance stores either an IPv4 address or an
    IPv6 address. \cppclass{IPvXAddress} is mainly used in the transport layer and above
    to abstract away network addresses. It can be assigned from both \cppclass{IPAddress}
    and \cppclass{IPv6Address}, and can also parse string representations of both.
    The \ffunc{isIPv6()}, \ffunc{get4()} and \ffunc{get6()} methods can be used
    to access the value.
\end{itemize}




\section{The Interface Table}

The \nedtype{InterfaceTable} module holds one of the key data structures in
the INET Framework: information about the network interfaces in the host.
The interface table module does not send or receive messages; other modules
access it using standard C++ member function calls.

\subsection{Accessing the Interface Table}

If a module wants to work with the interface table, first it needs to obtain a
pointer to it. This can be done with the help of the
\cppclass{InterfaceTableAccess} utility class:

\begin{cpp}
IInterfaceTable *ift = InterfaceTableAccess().get();
\end{cpp}

\cppclass{InterfaceTableAccess} requires the interface table module to be a
direct child of the host and be called \ttt{"interfaceTable"} in order to
be able to find it. The \ffunc{get()} method never returns \ttt{NULL}: if
it cannot find the interface table module or cannot cast it to the
appropriate C++ type (\cppclass{IInterfaceTable}), it throws an exception
and stop the simulation with an error message.

For completeness, \cppclass{InterfaceTableAccess} also has a
\ffunc{getIfExists()} method which can be used if the code does not require
the presence of the interface table. This method returns \ttt{NULL} if the
interface table cannot be found.

Note that the returned C++ type is \cppclass{IInterfaceTable}; the initial
"\ttt{I}" stands for "interface". \cppclass{IInterfaceTable} is an abstract
class interface that \cppclass{InterfaceTable} implements. Using the abstract
class interface allows one to transparently replace the interface table with
another implementation, without the need for any change or even
recompilation of the INET Framework.

\subsection{Interface Entries}

Interfaces in the interface table are represented with the
\cppclass{InterfaceEntry} class. \cppclass{IInterfaceTable} provides member
functions for adding, removing, enumerating and looking up interfaces.

Interfaces have unique names and interface IDs; either can be used to look up
an interface (IDs are naturally more efficient). Interface IDs are invariant to
the addition and removal of other interfaces.

Data stored by an interface entry include:

\begin{itemize}
  \item \textit{name} and \textit{interface ID} (as described above)
  \item \textit{MTU}: Maximum Transmission Unit, e.g. 1500 on Ethernet
  \item several flags:
    \begin{itemize}
      \item \textit{down}: current state (up or down)
      \item \textit{broadcast}: whether the interface supports broadcast
      \item \textit{multicast} whether the interface supports multicast
      \item \textit{pointToPoint}: whether the interface is point-to-point link
      \item \textit{loopback}: whether the interface is a loopback interface
    \end{itemize}
  \item \textit{datarate} in bit/s
  \item \textit{link-layer address} (for now, only IEEE 802 MAC addresses are supported)
  \item \textit{network-layer gate index}: which gate of the network layer within the host the NIC is connected to
  \item \textit{host gate IDs}: the IDs of the input and output gate of the host the NIC is connected to
\end{itemize}

Plus: network-layer protocol specific data: ipv4Data(), ipv6Data(), protocol3Data(), protocol4Data()

Extensibility:

by NICs: inheritance;

by network layer protocols: interfacedata like ipv4(), ipv6()


\subsection{Interface Registration}

Interfaces are registered dynamically in the
initialization phase by modules that represent network interface cards (NICs).

\subsection{Interface Change Notifications}

\subsection{Resolving addresses with IPAddressResolver}

\subsection{Auto address assignment by FlatNetworkConfigurator}




\section{The Routing Table}

\nedtype{RoutingTable}:
This module contains the IP (v4) routing table, and heavily relies
on \nedtype{InterfaceTable} for its operation. This module is also accessed from other
modules (typically IP) by calling the public member functions of its C++ class.
There are member functions for querying, adding, deleting routes, and
finding the best matching route for a given destination IP address.
The routing table submodule within the host (router) must be called
\ttt{routingTable} for other modules to find it.


TODO


\section{Initialization Stages}

Node architecture makes it necessary to use multi-stage initialization.
What happens in each stage is this:

In stage 0, interfaces register themselves in \nedtype{InterfaceTable} modules

In stage 1, routing files are read.

In stage 2, network configurators (e.g. \nedtype{FlatNetworkConfiguration})
assign addresses and set up routing tables.

In stage 3, TODO...

In stage 4, TODO...

...

The multi-stage initialization process itself is described in the OMNeT++ Manual.


\section{Communication between protocol layers}

In the INET Framework, when an upper-layer protocol wants to send a data
packet over a lower-layer protocol, the upper-layer module just sends the
message object representing the packet to the lower-layer module, which
will in turn encapsulate it and send it. The reverse process takes place
when a lower layer protocol receives a packet and sends it up after
decapsulation.

It is often necessary to convey extra information with the packet. For
example, when an application-layer module wants to send data over TCP, some
connection identifier needs to be specified for TCP. When TCP sends a
segment over IP, IP will need a destination address and possibly other
parameters like TTL. When IP sends a datagram to an Ethernet interface for
transmission, a destination MAC address must be specified. This extra
information is attached to the message object to as \textit{control info}.

Control info are small value objects, which are attached to packets
(message objects) with its \ttt{setControlInfo()} member function.
Control info only holds auxiliary information for the next protocol layer,
and is not supposed to be sent over the network to other hosts and routers.


\section{Publish-Subscribe Communication within Nodes}

The \nedtype{NotificationBoard} module makes it possible for several modules to
communicate in a publish-subscribe manner. For example, the radio module
(\nedtype{Ieee80211Radio}) fires a \textit{"radio state changed"} notification when
the state of the radio channel changes (from TRANSMIT to IDLE, for example),
and it will be delivered to other modules that have previously subscribed
to that notification category. The notification mechanism uses C++ functions
calls, no message sending is involved.

The notification board submodule within the host (router) must be called
\ttt{notificationBoard} for other modules to find it.


\section{Network interfaces}

todo...

\section{The wireless infrastructure}

todo...



\section{NED Conventions}

\subsection{The @node Property}

By convention, compound modules that implement network devices (hosts,
routers, switches, access points, base stations, etc.) are marked with the
\ttt{@node} NED property. As node models may themselves be hierarchical, the
\ttt{@node} property is used by protocol implementations and other simple
modules to determine which ancestor compound module represents the physical
network node they live in.

\subsection{The @labels Module Property}

The \ttt{@labels} property can be added to modules and gates, and it
allows the OMNeT++ graphical editor to provide better editing experience.
First we look at \ttt{@labels} as a module property.

\ttt{@labels(node)} has been added to all NED module types that may occur on
network level. When editing a network, the graphical editor will NED types
with \ttt{@labels(node)} to the top of the component palette, allowing the
user to find them easier.

Other labels can also be specified in the \ttt{@labels(...)} property. This
has the effect that if one module with a particular label has already been
added to the compound module being edited, other module types with the same
label are also brought to the top of the palette. For example,
\nedtype{EtherSwitch} is annotated with \ttt{@labels(node,ethernet-node)}.
When you drop an \nedtype{EtherSwitch} into a compound module, that will
bring \nedtype{EtherHost} (which is also tagged with the
\ttt{ethernet-node} label) to the top of the palette, making it easier to
find.

\begin{ned}
module EtherSwitch
{
    parameters:
        @node();
        @labels(node,ethernet-node);
        @display("i=device/switch");
    ...
}
\end{ned}

Module types that are already present in the compound module also appear in
the top part of the palette. The reason is that if you already added a
\nedtype{StandardHost}, for example, then you are likely to add more of the
same kind. Gate labels (see next section) also affect palette order: modules
which can be connected to modules already added to the compound module
will also be listed at the top of the palette. The final ordering is the
result of a scoring algorithm.


\subsection{The @labels Gate Property}

Gates can also be labelled with \ttt{@labels()}; the purpose is to make it easier
to connect modules in the editor. If you connect two modules in the editor,
the gate selection menu will list gate pairs that have a label in common.

TODO screenshot

For example, when connecting hosts and routers, the editor will offer connecting
Ethernet gates with Ethernet gates, and PPP gates with PPP gates. This is the
result of gate labelling like this:

\begin{ned}
module StandardHost
{
    ...
    gates:
        inout pppg[] @labels(PPPFrame-conn);
        inout ethg[] @labels(EtherFrame-conn);
    ...
}
\end{ned}

Guidelines for choosing gate label names: For gates of modules that
implement protocols, use the C++ class name of the packet or acompanying
control info (see later) associated with the gate, whichever applies;
append \ttt{/up} or \ttt{/down} to the name of the control info class. For
gates of network nodes, use the class names of packets (frames) that travel
on the corresponding link, with the \ttt{-conn} suffix. The suffix prevents
protocol-level modules to be promoted in the graphical editor palette when
a network is edited.

Examples:

\begin{ned}
simple TCP like ITCP
{
    ...
    gates:
        input appIn[] @labels(TCPCommand/down);
        output appOut[] @labels(TCPCommand/up);
        input ipIn @labels(TCPSegment,IPControlInfo/up);
        output ipOut @labels(TCPSegment,IPControlInfo/down);
        input ipv6In @labels(TCPSegment,IPv6ControlInfo/up);
        output ipv6Out @labels(TCPSegment,IPv6ControlInfo/down);
}
\end{ned}


\begin{ned}
simple PPP
{
    ...
    gates:
        input netwIn;
        output netwOut;
        inout phys @labels(PPPFrame);
}
\end{ned}



%%% Local Variables:
%%% mode: latex
%%% TeX-master: "usman"
%%% End:

